---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First, set and load the working environment and necessary libraries.

```{r, echo=TRUE, results=FALSE}
library(knitr)
library(xtable)
library(plyr)
library(dplyr)
setwd("C:/Users/rgrillo/datasciencecoursera/RepData_PeerAssessment1")
```

Load and preprocess the data now.

```{r, echo=TRUE}
activity <- read.csv(unzip(zipfile = "activity.zip", files = "activity.csv"))
```

Viewing the data separately, using the View() command (not written), reveals that the data needs no pre-processing.

## What is mean total number of steps taken per day?

There are 61 total days recorded.  We use ddply to calculate the total number of steps taken per day, and their mean and median.

```{r, echo=TRUE}
days <- length(unique(activity$date))
stepsPerDay <- ddply(activity, ~date, summarise, recorded_number_of_steps_taken = sum(steps, na.rm = TRUE))
stepsPerDay
hist(stepsPerDay[ , 2])
meanSteps <- mean(stepsPerDay[ , 2], na.rm = TRUE)
medianSteps <- median(stepsPerDay[ , 2], na.rm = TRUE)
```

The mean number of steps taken per day is `r meanSteps` and the median number of steps taken per day is `r medianSteps`.

## What is the average daily activity pattern?

The average number of steps taken at each five-minute interval is plotted below.

```{r, echo=TRUE}
avgDailyActivity <- ddply(activity, ~interval, summarise, average_number_of_steps_taken = mean(steps, na.rm = TRUE))
avgDailyActivity
plot(x = avgDailyActivity[,1], y = avgDailyActivity[ , 2], type = "l")
maxIntervalIdx <- which(avgDailyActivity[ , 2] == max(avgDailyActivity[ , 2]))
maxInterval <- avgDailyActivity[maxIntervalIdx, 1]
time <- maxInterval/60
time2 <- paste(floor(time), round((time-floor(time))*60), sep=":")
```

The five-minute interval that contains the maximum number of steps across all days, on average, is `r maxInterval`, which is `r time2`.

## Imputing missing values

```{r, echo=TRUE}
numNA <- length(which(is.na(activity$steps)))
```

There are `r numNA` cases with missing values in the dataset.  We will use the mean for the 5-minute interval to fill in the NAs.

```{r, echo=TRUE}
NewActivity <- activity
for(i in 1:nrow(NewActivity)) {
    if(is.na(NewActivity[i, 1])) {
        tmpIdx <- which(avgDailyActivity[ , 1] == NewActivity[i, 3])
        NewActivity[i, 1] <- avgDailyActivity[tmpIdx, 2]
    }
}
View(NewActivity)
```

Now we make a histogram of the data

```{r, echo=TRUE}
newStepsPerDay <- ddply(NewActivity, ~date, summarise, recorded_number_of_steps_taken = sum(steps))
newStepsPerDay
hist(newStepsPerDay[ , 2])
newMeanSteps <- mean(newStepsPerDay[ , 2], na.rm = TRUE)
newMedianSteps <- median(newStepsPerDay[ , 2], na.rm = TRUE)
diffMeanSteps <- abs(newMeanSteps - meanSteps)
diffMedianSteps <- abs(newMedianSteps - medianSteps)
```

The mean total number of steps taken per day is `r newMeanSteps` and the median total number of steps taken per day is `r newMedianSteps`, which is a difference of `r diffMeanSteps` and `r diffMedianSteps` for the respective values, computed without imputing new values for the NAs (having removed them previously). 

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
NewActivity$date <- as.POSIXct(NewActivity$date)
NewActivity$weekday <- weekdays(NewActivity$date)
if(NewActivity$weekday %in% c("Saturday", "Sunday")) {
    NewActivity$weekday_type <- "Weekend"
} else {
    NewActivity$weekday_type <- "Weekday"
}
```
